---
title: "Hierarchical Distance Sampling to Estimate Animal Abundance in Prey Lang Wildlife Sanctuary"
author: "Nan Nourn"
date: "September 19 2024"
  
bibliography: plws-ref.bib
csl: global-ecology-and-conservation.csl
  
toc: true
toc-depth: 3
  
comments:
  hypothesis: true

format:
  html:
    theme: minty
    number-sections: false
    monofont: Fira Code
    code-fold: show
    code-link: true
    code-overflow: wrap
    downloads: [docx, epub, pdf]
    sharing: [twitter, facebook]
#  pdf:quarto
#    documentclass: scrreprt
editor: visual
---
# Packages
```{r}
#| label: library
#| message: false
#| warning: false
library(tidyverse)
library(sf)
library(tmap)
library(here)
```

# Abstract 

This chapter explores distance sampling (DS) analyses to estimate wildlife abundance and density in Cambodia. Specifically, we attempt to estimate Great hornbill (GH)*Buceros bicornis* density within Prey Lang Wildlife Sanctuary (PLWS) in central Cambodia utilizing occurrence and recorded distance data collected from line transect field surveys.

Conservation International-Cambodia (CI) manages a REDD+ project within protected area Prey Lang Wildlife Sanctuary, located in central Cambodia. REDD+ programs require the assessment of biodiversity monitoring to determine the impact of carbon conservation measures and its impact on native flora and fauna species. In accordance to this REDD+ project within PLWS, CI deployed a first assessment of transect line surveys to assess baseline activity of key biodiversity species inhabiting PLWS.

Specifically, within the DS models, our objectives were to explore how the use of hierarchical distance sampling methods utilizing covariates such as GH group size, and other environmental covariates, such as tree canopy height, affect densities for GH within PLWS. Ultimately, we provide the first robust density and population estimates for GH for Cambodia. 

# Introduction

## REDD+

Global deforestation in tropical forests has led to continued decline of wildlife populations and increases in forest habitat degradation, contributing to the loss of critical ecosystem functioning and biodiversity decline leading affectiving important ecosystem processes. This is especially prominent in Global South countries, where human population growth is on the rise, leading to an increase in middle class economies.  To keep up with the demand for human development, global south governments must keep up with with the growth of industrialization and development while allowing for sustainable outcomes when conserving ecologically important ecosystems.

**Reduced Emissions from Deforestation and Forest Degradation (REDD+) is a voluntary global scale climate change mitigation program aiming at creating financial values for carbon stored in forests, and aimed at offerring incentives to reduce greenhouse gas emissions from deforestation and forest degradation** [@WorldBank 2013].

In many cases, REDD+ provides a complementary avenue for forest protection by providing financial-based incentives as conservation projects thru reduced forest loss against a defined baseline [@pauly2022sr]. Through this mechanism, REDD+ proponents have supported the notion that Global North nations can contribute to sustainable development in Global South naitons by investing in these carbon markets created through theoretical protection of forests. 

In an effort to curtail forest loss and combat deforestation, REDD+ program schemes have been implemented globally in developing tropical regions along with generating carbon credits to be used.

## Cambodia

Cambodia is considered a global biodiversity hotspot [@myers2000n]. Within Southeast Asia, Cambodia holds the largest contiguous remaining tropical forest habitat within the region, containing important habitat types such as evergreen, semi-evergreen and deciduous forests, along with savanna grasslands.

However, Cambodia is plagued with deforestation problems due to corruption in the government and a lack of enforcement to safeguard natural resources.  Government officials are likely to benefit from selling of land concessions and forest trees as logs. Cambodia is also a developing country meeting high demand for a growing population, such as the need to expand infrastructure to support civil society. The main drivers of biodiversity decline is attributed through habitat loss from deforestation along with illegal poaching of wildlife for trade markets. 

Efforts to protect remaining tropical forests have begun seeing Cambodia adopting and initializing REDD+ programs in protected areas, usually in collaboration with the Cambodian Royal government, conservation non-government organizations and private companies.

In an an effort to generate carbon credits for sale, the JCM REDD+ program in Prey Lang Wildlife Sanctuary (PLWS) was initiated to ensure sustainable financing for the program. In 2018 to 2021, the JCM REDD+ Project Phase 1 was implemented in Stung Treng within PLWS, while Phase 2 project proposal was signed by Cambodia's Ministry of Evironment (MoE), Conservation International (CI), and Mitsui & Co., Ltd., (Mitsui) in April 2021 to extend the project objectives to include the other three provices of Kampong Thom, Kratie, and Preah Vihear, with a completed end date of 2026. 

In collaboration with the Cambodian Ministory of Environment (MoE), Conservation International (CI) has had a long-term partnership with the goal to conserve tropical forest natural resources. Various conservation programs such as protected area management, biodiversity monitoring and community livelihood improvements have been implemented to assess drivers of deforestation and other illegal activities regarding natural resources in PLWS.

## Great Hornbills

Hornbills are large forest birds classified in the order of Bucerotiformes [@corlett2017geaca] consisting of 15 genera, 62 species; with 32 species inhabiting Asia and 25 species in Africa [@poonswad2013cb]. Prominent characteristics are an oversized, long curved bill with a casque on top. Hornbills are known as the most important single family of seed dispersal agents for large-woody species in the forest canopy [@corlett2017geaca].

Greater hornbill (GH) *Buceros bicornis* can be found in Cambodia occurring through their natural habitat of dense tropical evergreen forest.

## Distance Sampling to Estimate Abundance and/or Density

Distance sampling (DS) has been a popular type of statistical modelling used to analyze and estimate abundances and/or densities of wildlife populations. DS has often served as effective models to estimate wildlife population numbers for wildlife management quotas and has also been used for conservation measure and biodiversity assessments to estimate biodiversity goals.

The main assumptions for classic DS models include utilizing recorded distances from the animal as a detection function.

Here, we estimate a species to account for both ecological and observation processes.

### Hierarchical Distance Sampling (HDS)

As noted before, traditional distance sampling involves pooling all distance data from all $S$ spatial locations and estimating the parameter(s) of the detection function, $\sigma$ for the half-normal. In turn this is used to obtain an esitmate of density, witht he variance based ont he encounter rate variance, which pulls infoormation among the sample units. However, we may be more interested in modeling variaiton in $N_s$ or local density across sample units $s$ or making explicit predictions at other transects or point locations. 
#### Previous Hornbill Studies Utilizing Distance Sampling

A number of peer-reviewed studies have incorporated line transect surveys and distance sampling (DS) analyses to estimate hornbill densities from different regions of the world. 

@sriprasertsil2024geac conducted variable-width line transect surveys and used multi-covariate distance sampling to estimate hornbill densities from June 2021 to June 2022 in two protected areas in Southern Thailand: Hala-Bala Wildlife Sanctuary and Budo Su-Ngai Padi National Park. Transect lines were monitored 8-12 times across these 13 months. Authors also conducted fruit tree watches in Bala and Budo Mountain for hornbill visitation between August 2022 and May 2023 following established methods (Naniwadekar et al. 2019). 

@pradhan2024geac estimated abundances for four hornbill species: Great Hornbill, Wreathed Hornbill, Rufous-necked Hornbill, Oriental Pied-Hornbill) in the highly-fragmented and highly-modified Buxa Tiger Reserve in northern West Bengal, India. Researchers used line transects and multi-covariate distance sampling to estimate densities recording hornbill species, flock size, time of sighting, activity (perched, foraging, call and flight) and perpendicular distance from the observer to the bird. Researchers walked transects over two non-breeding seasons of September 2019 to March 2020 and November 2020 to March 2021. Predictor variables (elevation, density of fruiting trees, cut logs, total basal area) and hornbill occurrence was explored using a generalized linear model with binomial error structure with all hornbill occurrences pooled together. Density estimates for Great Hornbill, Wreathed Hornbill and Rufous-necked Hornbills were very low (\< 1 per km\^2), whereas Oriental Pied-Hornbill was 10.6 per km\^2. Authors recommend that even with spareness of detections of larger-bodied species, they were still able to obtain reliable density estimates by using body size as a categorical covariate for modelling the detection function.

@mynott2021ara conducted line transect and point count surveys in the Phillipines to understand endemic Visayan (or Tarictic) Hornbill *Penelopides panini* densities.  24 randomized grid cells were surveyed once between JAnuary through March 2020. Within each cell, \\ 1 to 4 transects of 0.3 to 2.5 km were positioned randomly or along existing narrow trails. Distance sampling point counts were used to survey hornbills along transects. Point counts were situated \> 200 m apart from transects to minimize likelihood of recording the same birds from multiple points. Count periods were 10 minutes long. Estimated hornbill densities were significantly higher in primary forest habitat (17.8/km\^2) than secondary forest habitat (3.7/km\^2) Model averaging identified two variables as significant predictors of hornbill presence: distance from park boundary (positive) and the interaction between medium-sized trees and distance from the Park's boundary (negative). Density estimates reveal forest cover with minimal disturbance is important.

@pawar2021i utilized distance sampling methods to estimate abundance of Great Hornbills and Malabar Grey Hornbills in two contrasting habitat types in protected area rainforests versus fragmented habitat consisting of coffee plantations. They found that both hornbill species used both habitat types year-round. Their main findings include determining that distance sampling estimates were higher in PA for both nesting and non-nesting seasons.

@naniwadekar2013tcs estimated hornbill densities by sampling in four sessions (years) from 2009-2012 with a large sampling effort of 842.1 km walked in Namdapha Tiger Reserve in India. They sampled mainly during the non-breeding season across two phases. All data was pooled together due to having less than 40 detections in each of the strata in the two analyses.

@gale2006bci used variable-width line transect surveys to estimate densities of seven hornbill species. They surveyed 11 transects once per month totaling 68 km in length from January 2001 to April 2002. Transects ranged from 3 to approximately 11.4 km in length. Density estimates were obtained from DISTANCE program with all birds entered as clusters and distance data grouped automatically by the software, combining both aural and visual observations.

@raman2003bci used data from line transects to compare with Malabar Grey Hornbill point-count density estimates from Karian Shola Naitonal Park in Anamalais, India. Data were analyzed in program DISTANCE with the Half-Normal model based on minimum AIC value. One of the first studies to employ systematic line transect surveys.

@datta1998o conducted line transect surveys across various habitat types during December 1995 to April 1996. Distance sampling analysis was not used for this study.

# Methods

## Study Area: Prey Lang Wildlife Sanctuary (PLWS)

Prey Lang Wildlife Sanctuary (PLWS) is a protected area located in the central plains of Cambodia. As one of the largest areas of contiguous lowland tropical forests left in mainland Southeast Asia, PLWS was established in 2016 and covers four provinces including 1) Kampong Thom, 2) Preah Vihear, 3) Kratie and 4) Stung Treng. 

PLWS is notable for its immense biodiversity, providing critical refuge for up to 55 globally threatened vertebrates including gibbons, Asian elephants, and /insert carnivore/, while having up to 538 plant and tree species. PLWS is also critical in that it supports more than 250,000 people inhabiting the forest, including the indigenous Kuy people who reside in and around this protected area.  PLWS also serves as an important watershed for the Mekong and the Tonle Sap lake.

While maintaining protected area status, PLWS continues to face enormous natural resource extractive pressure due to weak enforcement of environmental laws and lack of resources. Illegal forest logging, wildlife trade, land encroachment, and unsustainable agricultural practices have led to continued degradation of quality forest habitat.

PLWS is comprised mainly of evergreen and semi-evergreen forest, mixed with some deciduous forests and savanna grasslands. 

## Line Transects

Transect grids/lines, which could also be noted as "survey sites" were generated in ArcMap 10.8.1 and placed randomly within PLWS outside of farms, degraded areas, human settlement, and other non-forest habitat. Transect grids were composed of four 1-km lines forming a square grid, totaling 4 km in walked transect length (1 km^2). 

Before data collection, transect lines were cleared using machetes to create a trail no more than one meter in width to ensure minimum impacts on local vegetaion. Trails were painted blue every 10 to 20 meters, while transect grid corners were painted by applying paint on trees. Field training and practice trials were conducted before implementation of data collection.

Fifteen key biodiversity species were selected and targeted for biodiversity monitoring at PLWS. 

Data collection occurred during January to May 2023 during Cambodia's dry season. Transect grids were mainly completed in evergreen and semi-evergreen forest, with a few transect lines conducted in dry deciduous forest habitat. Each transect grid was surveyed twice per day, with a morning and afternoon session.

Here is a map of transects:
```{r}
#| label: plws-map
#| warning: false
#| message: false
# load data
plws_transect <- read_csv(here("data","2023-plws-line-transect-data-nn.csv"))
#plws_transect |> glimpse()
# extract transect grid data 
grid <- plws_transect |> 
  select(grid_id, grid_line, grid_line_direction, 
         x_mid, y_mid, x_transect, y_transect, province) |> 
  drop_na() |> 
  st_as_sf(coords=c("x_transect","y_transect"), crs=32648)
# extract midpoint grid data
grid_midpt <- grid |> 
  group_by(grid_id) |>
  distinct(x_mid, y_mid) |> 
  ungroup() |> 
  st_as_sf(coords=c("x_mid","y_mid"), crs=32648)
# create grid polygons
grid_polygon <- grid |> 
  group_by(grid_id) |> 
  summarise(do_union=FALSE) |>  
  st_cast("LINESTRING")
# map
tmap_mode("view")
tm_shape(grid) +
  tm_dots(col="grid_line_direction", size = 0.075) +
  tm_text("grid_line", size=0.7) +
  tm_shape(grid_midpt) +
  tm_text("grid_id", col="red", size=1) +
  tm_shape(grid_polygon) +
  tm_lines()
```

# Results

# Discussion

# Conclusions

## Management Implications

# References

::: {#refs}
:::
